\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-20 rebuttal.tex

\begin{document}

\signature{
	Arne Van Den Kerchove \\
	\texttt{arne.vandenkerchove@kuleuven.be} \\
	+32 473 32 78 71 \\
}
\address{
	KU Leuven \\
	Laboratory for Neuro- and Psychophysiology \\
	Campus Gasthuisberg, O\&N 2 \\
	Herestraat 49 bus 1021 \\
	BE-3000 Leuven
}

\begin{letter}{
	prof. John Creemers \\
	KU Leuven \\
	Doctoral School of Biomedical Sciences \\
	Campus Gasthuisberg, O\&N 4 \\
	Herestraat 49 bus 700
	BE-3000 Leuven

}
\opening{Dear chair and members of the reading committee}

I would like to extend my sincere gratitude for the time and effort you have taken
to review the manuscript of my PhD dissertation.
Additionally, I would like to thank you for the insightful, constructive and
detailed comments you all have provided.
The following reply contains a point-by-point response to your review comments.
Attached, you will also find an updated draft of the manuscript, as well as a
version with indicated changes for your approval.

@@

\reply{Andrea K\"ubler}
Major comments:
\begin{enumerate}
	\item
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item
\end{enumerate}

\reply{Fabien Lotte}
Major comments:
\begin{enumerate}
	\item Part of the confusion between active and passive BCI might be due
	to the poor choice of phrasing ``paradigms that require little active
  participation from the user, like redirecting attention or initiating imagined actions, are classified as passive BCI.''.
	This has been rewritten to ``in contrast to paradigms that require the user's active participation, for
	instance through redirecting attention or initiating imagined actions,
	paradigms free of this requirement are classified as passive bci,''
	for clarity.
	Neurofeedback has been reclassified as active BCI in the text and in
	figure 1.3.
	\item \todo{Why XDAWN+RG no shrinkage (performed worse), fair
	comparison of sample rate}
  \item Appendix 1.A now fully reports all obtained accuracies to give a better
    insight in effect sizes. Wherever the significant effects have been mentioned
    in the text, the effect size has now been reported along with it.
	\item In the results section of chapter 4, statistical comparisons were
	only performed between methods for which we evaluated the performance
	ourselves (HODA, PARAFACDA and BTTDA).
	Since the MOABB benchmarking framework~\cite{Aristimunha2023} allows for
	fair comparison of performance metrics reported across different works,
	the analyses for other comparison methods were not re-run
	by us as reported, but performances were
	taken from the MOABB benchmark database~\cite{Chevallier2024}.
	However, this source provided only the aggregated performance metrics
	and not the metrics per subject, session and cross-validation fold,
	hence we could not perform relevant statistical comparison with these
	comparison models.
\item Tables 6.2 and 6.3 now report the absolute cross-condition ROC-AUC values
  per model and dataset for a more complete and interpretable presentation of
    the results, and the absolute values have been mentioned when referring to
    effects in the results section.
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item According to the annotations, typographical errors have been
	corrected in text and figures, and imprecise or inaccurate statements have been adjusted.
	The recommended references have been discussed and, where necessary,
	statements were updated accordingly.
  Figure 1.5 was corrected and matched with the correct caption.
\end{enumerate}

\reply{Reinhold Scherer}
Major comments:
\begin{enumerate}
	\item \todo{quantify qualitative statements} Throughout the text, qualitative
    statements have been qualified where possible and appropriate.
	\item  All mathematical equations and derivations have thoroughly
	been checked for correctness and notational consistency, and have been
	corrected where necessary.
	\item As indicated by the reviewer, true on-line experiments with live
	feedback have indeed not reported in this thesis, but were foreseen to take
	place within the framework of this doctoral study.
	I agree that these experiments would be valuable to supplement the
	presented conclusions and are the logical next step.
	Indeed, the start of experimentation with healthy control participants,
	as well as with patient groups, was heavily delayed by the COVID-19 pandemic, as the
	preparatory research started in September 2019 and I officially
	enrolled in the doctoral training program in January 2020.
	This restricted not only our access to experiment subjects for a
	significant period at the start of the PhD, and even longer for
	vulnerable patient groups, but also shifted priorities for the medical
	staff at our partner institutions and patient centers.
	Hence, work early in the PhD mainly focused on the development of
	computational methods.
	We are currently starting up true on-line experimentation in cooperation
	with our partner patient centers, with an adapted interface and experimental protocol.
  This limitation has now clearly been mentioned in the future work section of
  chapter 8.
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item Recorded datasets could not be made publicly available since the
	ethical protocol (Ethics Commission of University Hospital Leuven --
	S62547 v1.4) for the conducted studies did not include provisions
	publishing the data.
\end{enumerate}

\reply{Maarten De Vos}
Major comments:
\begin{enumerate}
	\item Mathematical equations and derivations have thoroughly
	been checked for correctness and notational consistency, and have been
	corrected throughout the manuscript where necessary.
	The paragraph on Block-Toeplitz linear discriminant analysis (tLDA) in
	section 1.6.4 has also been corrected, and it was expanded for clarity.
	Figures 3.1 and 3.2 have been regenerated and relabeled for consistent
	formatting.
	\item The Wilcoxon signed-rank, a non-parametric,
    paired test was used to compare ROC-AUC values in sections 4.3.3 and 6.3.1.
    I agree that the DeLong test is indeed an appropriate test for
    statistically comparing ROC-AUCs between two models, but it is commonly used
    to compare ROCs obtained from a single test dataset (potentially obtained
    through stacking cross-validated predictions).
    Our experiments, on the other hand, involve calculating ROC-AUCs per
    cross-validation fold, yielding datasets of multiple ROC-AUCs to compare.
    In this case, the Wilcoxon signed-rank test is the appropriate
    test~\cite{Rainio2024}.
    While both approaches are valid strategies and commonly used, I believe
    the latter is more interpretable in the context of the generalizability
    of machine learning models, since it more directly reflects the variance
    across folds.
    Hence, I recognize the potential application of the DeLong test here,
    but I would argue that using Wilcoxon signed-rank tests in combination with
    the evaluation strategy employed in the experiments is still a valid approach.
   \end{enumerate}
Minor comments:
\begin{enumerate}
	\item Unsupported claims regarding generalisability have been omitted
	from the abstract.
	\item Throughout the manuscript, different EEG band-pass filtering
	settings have been used.
	Chapter 1 indicates that ERPs are usually band-pass filtered between
	0.5 and 16 Hz.
	These were also the preferred cut-off frequencies for other
	(unpublished) analyses used during the course of the thesis work, an
	those used for the analyses in chapter 3.
	However, ERP datasets evaluated in chapter 4 were band-pass filtered
	between 1 and 24 Hz, and in chapter 6 between 0.1 and 20 Hz.
	While there is no general consensus in the field about the optimal
	cut-off frequencies for P300 detection and ERP analysis in general,
	these values all fall well within the accepted
	bounds~\cite{Bougrain2012}.
	Decisions to deviate from this were made consciously.
	In chapter 3, filtering between 1 and 24 Hz was applied for consistency
	with the recommended ERP preprocessing pipeline of the MOABB
	benchmarking framework~\cite{Aristimunha2023}.
	This allows for a fair comparison under equal circumstances of the
	performance of the proposed algorithm with performances of other
	algorithms reported by the MOABB benchmarking
	database~\cite{Chevallier2024}.
	In chapter 6, the low-pass frequency was increased from 16 Hz to 20 Hz,
	since the analyses in this explicitly involve the short-time effects of
	latencies of ERP components, and the analysis of faster, early ERP
	components. A low low-pass frequency would harm the precision of
	latency effects in the filtered signal.
	A lower high-pass filter of 0.1 Hz was used here to reveal the
	influence of the proposed alignment approach on potential
	slower late components.
	\item Typographical errors have been corrected throughout the
	manuscript.
\end{enumerate}

\reply{Adalberto Simeone}
Major comments:
\begin{enumerate}
	\item \todo{discuss in limitations the impact of letters and future
	work experiments with an actual usable interface}
	\todo{mention here plans to go to patients with actual speller
	interface}
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item Measurement unit representation has been uniformized according to the
    SI standard throughout the text and in figures 5.4-5.6.
  \item Numerical citations have now consistently been replaced by text-style citations
    mentioning the author names where appropriate in the sentence.
  \item Contractions have been remove throughout the text.
\end{enumerate}

### 2024-11-20 chapters/bci.tex
Reply to DeLong test

\chapter{Brain-computer interfaces}
\label{sec:bci}
\epigraph{``My mind controls my reality.''}{Frank Herbert -- Children of Dune}

@@

Starting in the 1970s from methods to establish basic communication from minimal brain
activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
into sophisticated interaction schemes that restore or enhance many capabilities by
bypassing defective parts of the nervous or muscle system.\todo{the BCI concetp
was proposed at that time, but maybe not yet this form of communication. Also,
this reference is not from the 1970s, you may want to cite the original paper}
By now, classic \acp{bci} methods are starting to mature outside the lab setting and can
be used as assistive communication technology by individuals with \acl{sspi}
~\cite{Wolpaw2018,Milekovic2018}.
Impressive cutting-edge experimental examples include a brain-spine interface allowing a
paraplegic individual to walk again~\cite{Lorach2023}, a \ac{bci} translating internal
speech to a facial avatar speaking the imagined text~\cite{Metzger2023}, and fast
communication through decoding imagined handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

>>>>

\chapter{Brain-computer interfaces}
\label{sec:bci}
\epigraph{``My mind controls my reality.''}{\emph{Children of Dune -- Frank
Herbert}}

@@

\Acp{bci} Started as a speculative concept in the 1970's~\cite{Vidal1973}.
Now, they have gradually evolved from methods  developed in the the 1990's to
establish basic communication from minimal brain activity
(yes/no, left/right, \ldots)~\cite{Wolpaw2002} into sophisticated interaction
schemes that restore or enhance many capabilities by bypassing defective parts
of the nervous or muscle system.
By now, classic \acp{bci} methods are starting to mature outside the lab setting and can
be used as assistive communication technology by individuals with \acl{sspi}
~\cite{Wolpaw2018,Milekovic2018}.
Impressive cutting-edge experimental examples include a brain-spine interface allowing a
paraplegic individual to walk again~\cite{Lorach2023}, a \ac{bci} translating internal
speech to a facial avatar speaking the imagined text~\cite{Metzger2023}, and fast
communication through decoding imagined handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

### 2024-11-20 chapters/conflict.tex
Self-plagiarism check
Reply to DeLong test

Arne Van Den Kerchove is funded by a bilateral cooperation program between KU
Leuven (special research fund of the KU Leuven GPUDL/20/031) and the university
of Lille under the Global PhD Partnership program.
Prof. Marc Van Hulle is supported by research grants received from
the European Union’s Horizon Europe Marie Sklodowska-Curie Action program
(grant agreement No. 101118964), the European Union’s Horizon 2020 research and
innovation program (grant agreement No. 857375), the special research fund of
the KU Leuven (C24/18/098), the Belgian Fund for Scientific Research – Flanders
(G0A4118N, G0A4321N, G0C1522N), and the Hercules
Foundation (AKUL 043).

>>>>

Arne Van Den Kerchove is funded by a bilateral cooperation program between KU
Leuven (special research fund of the KU Leuven GPUDL/20/031) and the University
of Lille under the Global PhD Partnership program.

Prof. Marc Van Hulle is supported by research grants received from
the European Union’s Horizon Europe Marie Sklodowska-Curie Action program
(grant agreement No. 101118964), the European Union’s Horizon 2020 research and
innovation program (grant agreement No. 857375), the special research fund of
the KU Leuven (C24/18/098), the Belgian Fund for Scientific Research – Flanders
(G0A4118N, G0A4321N, G0C1522N), and the Hercules
Foundation (AKUL 043).

The authors also acknowledge the support by the RITMEA project co-financed by the
European Union with the European Regional Development Fund, the French state,
and the Hauts-de-France Region Council.
